Synthesizing Quotient Lenses

### 2018-03-14 papers/QuotientLenses/main.tex
project to collapse

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkred}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens,let,get,put,rquot,lquot},
emphstyle=\upshape\sffamily\color{dkred}, 
columns=fullflexible,
escapechar=\#,
mathescape, 
xleftmargin=1.5em,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
}
\makeatletter
     \let\lst@oldvisiblespace\lst@visiblespace
     \def\lst@visiblespace{\,\lst@oldvisiblespace\,}
\makeatother

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\textcolor{dkred}{\ensuremath{\mathsf{#1}}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

For example, consider the following QRE for writing author names:
%
\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

@@

In this subsection, we develop a QRE specification of \bibtex{}
records, introducing various QRE combinators along the way. 
Our first step in this process is to define a
whitespace format, which externally matches any non-zero number of whitespace
characters. It converts any such whitespace into a single space
character, its canonical form. 
We use the QRE \kw{project} primitive to define this
whitespace-normalizing QRE. \bcp{Why is it called
``project''?} \ksf{select seems like it might be a better name as we
are choosing one representative element from a set of possibilities.}\bcp{Or
``collapse''?}

\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}\bcp{Done, I think.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

@@

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify 
in an optional \kw{with} clause that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp) $\to$ ",\n")
\end{lstlisting}

@@

For instance, assume that $f(s)=$ ``\textvisiblespace'' (a space
character) for all whitespace strings $s$. Then the 
project QRE \lstinline{wsp_sp} defined above can
be expressed using the normalize combinator as
\lstinline{normalize (wsp," ", f)}.

Figure~\ref{fig:example-qre} gives a QRE specification for the simple
\bibtex{} records we consider here.
\begin{figure}
\begin{lstlisting}
let wsp = [#\,\textvisiblespace\,#\n\t\r]+
let wsp_sp = project wsp $\to$ " "
let last_name = [A-Z][a-z]
let first_name = [A-Z][a-z]

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records.\ksf{In what sense is this a QRE
spec vs a QRE lens spec?}}\saz{This defines only one endpoint of the eventual
\bibtex{}--EndNote lens.  Maybe we should say ``definition'' instead of ``specification''?}
\label{fig:example-qre}
\end{figure}

@@

\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\newcommand{\bsep}{\ \ \sep{} \ \ }
\begin{align*}
Q := \; & \normalize{R_1}{R_2}{f} \bsep{} \id(R) \bsep \project{R}{s} \bsep \squash{Q_1}{Q_2}{f} \\
 | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1
\sep Q_2) \bsep Q^*
\end{align*}
where $Q$ ranges over QREs, 
$R$ ranges over regular expressions, 
$f$ ranges over functions between regular languages, and 
$s$ ranges over character strings.

@@

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q_1}{Q_2}{f} & W(Q_1) \sep W(Q_2) & K(Q_2) \\
\normalize{R_1}{R_2}{f} & R_1 & R_2 \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot\; Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep \; Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(Q_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(Q_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions for QRE combinators. 
In describing regular expressions, we use the notations $\sep$ and $\bigcup$ for
alternation, the notation $\cdot$ for concatenation, and the notation
$*$ for Kleene closure.  
We use the notation $S_n$ to denote the set of all permutations of the
numbers $1$ to $n$.
}
\label{fig:wk}
\end{figure}
The \kw{squash} and \kw{permutation} combinators have the two most
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$ because the \kw{squash} combinator
merges the whole language $W(Q_1)$ of $Q_1$ with the whole language $W(Q_2)$ of $Q_2$.
The function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonize(Q_2)$.

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\project{R}{s})(w) &=& s\\[1ex]
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\[3ex]
\canonize(\normalize{R_1}{R_2}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\[1ex]
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\[3ex]
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers. \ksf{Solomon: please define/explain the . and *
operators on functions.}}
\label{fig:canonizers}
\end{figure}
The \kw{permutation} combinator gives rise to the most interesting
definition:
\[
\begin{array}{cl}
 & \canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= & (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{array}
\]
\noindent 
which places the strings that match $Q_i$ and $Q$ according to the
canonical permutation (i.e., the identity permutation) before applying
the canonizers $\canonize(Q_i)$ and $\canonize(Q)$, respectively.

@@

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_{\normalize{R_1}{R_2}{f}} \; w' &\iff&
f(w)=f(w') \\
w \; \equiv_{\id(R)} \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' &\iff& \mathit{True} \\ % \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well-Formed QREs}
\label{subsec:well-formed-qres}

@@

Strong unambiguity is necessary in the premises of the regular QRE combinators
because it ensures that the canonizing function of a QRE is well-defined.
For example, consider the QRE \lstinline{"a"* . (project "a"* $\to$ "a")}. The
behaviour of this QRE is not well-defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. We also require that the regular
combinators applied to the kernels of QREs be unambiguous since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.

@@

Figure~\ref{fig:qrerules} gives the inference rules for deriving well-formed
QREs.
\begin{figure}[tp!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\RightLabel{(Id)}
\UnaryInfC{$\wf{\kw{id}(R)}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\RightLabel{(Project)}
\UnaryInfC{$\wf{\project{R}{s}}$}
\DisplayProof
\hskip 1.5em
\end{center}

>>>>

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkred}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,collapse,perm,squash,normalize,using,ins,del,lens,let,get,put,rquot,lquot},
emphstyle=\upshape\sffamily\color{dkred}, 
columns=fullflexible,
escapechar=\#,
mathescape, 
xleftmargin=1.5em,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
}
\makeatletter
     \let\lst@oldvisiblespace\lst@visiblespace
     \def\lst@visiblespace{\,\lst@oldvisiblespace\,}
\makeatother

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\textcolor{dkred}{\ensuremath{\mathsf{#1}}}}
\newcommand{\collapse}[2]{\ensuremath{\kw{collapse} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

For example, consider the following QRE for writing author names:
%
\begin{lstlisting}
let wsp_sp = collapse wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

@@

In this subsection, we develop a QRE specification of \bibtex{}
records, introducing various QRE combinators along the way. 
Our first step in this process is to define a
whitespace format, which externally matches any non-zero number of whitespace
characters. It converts any such whitespace into a single space
character, its canonical form. 
We use the QRE \kw{collapse} primitive to define this
whitespace-normalizing QRE.

\begin{lstlisting}
let wsp_sp = collapse wsp $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}\bcp{Done, I think.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting}
\fi

@@

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify 
in an optional \kw{with} clause that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (collapse ("," . wsp) $\to$ ",\n")
\end{lstlisting}

@@

For instance, assume that $f(s)=$ ``\textvisiblespace'' (a space
character) for all whitespace strings $s$. Then the 
collapse QRE \lstinline{wsp_sp} defined above can
be expressed using the normalize combinator as
\lstinline{normalize (wsp," ", f)}.

Figure~\ref{fig:example-qre} gives a QRE specification for the simple
\bibtex{} records we consider here.
\begin{figure}
\begin{lstlisting}
let wsp = [#\,\textvisiblespace\,#\n\t\r]+
let wsp_sp = collapse wsp $\to$ " "
let last_name = [A-Z][a-z]
let first_name = [A-Z][a-z]

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (collapse ("," . wsp) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records.\ksf{In what sense is this a QRE
spec vs a QRE lens spec?}}\saz{This defines only one endpoint of the eventual
\bibtex{}--EndNote lens.  Maybe we should say ``definition'' instead of ``specification''?}
\label{fig:example-qre}
\end{figure}

@@

\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\newcommand{\bsep}{\ \ \sep{} \ \ }
\begin{align*}
Q := \; & \normalize{R_1}{R_2}{f} \bsep{} \id(R) \bsep \collapse{R}{s} \bsep \squash{Q_1}{Q_2}{f} \\
 | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1
\sep Q_2) \bsep Q^*
\end{align*}
where $Q$ ranges over QREs, 
$R$ ranges over regular expressions, 
$f$ ranges over functions between regular languages, and 
$s$ ranges over character strings.

@@

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\collapse{R}{s} & R & s \\
\squash{Q_1}{Q_2}{f} & W(Q_1) \sep W(Q_2) & K(Q_2) \\
\normalize{R_1}{R_2}{f} & R_1 & R_2 \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot\; Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep \; Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(Q_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(Q_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions for QRE combinators. 
In describing regular expressions, we use the notations $\sep$ and $\bigcup$ for
alternation, the notation $\cdot$ for concatenation, and the notation
$*$ for Kleene closure.  
We use the notation $S_n$ to denote the set of all permutations of the
numbers $1$ to $n$.
}
\label{fig:wk}
\end{figure}
The \kw{squash} and \kw{permutation} combinators have the two most
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$ because the \kw{squash} combinator
merges the whole language $W(Q_1)$ of $Q_1$ with the whole language $W(Q_2)$ of $Q_2$.
The function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonize(Q_2)$.

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\collapse{R}{s})(w) &=& s\\[1ex]
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\[3ex]
\canonize(\normalize{R_1}{R_2}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\[1ex]
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\[3ex]
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers. \ksf{Solomon: please define/explain the . and *
operators on functions.}}
\label{fig:canonizers}
\end{figure}
The \kw{permutation} combinator gives rise to the most interesting
definition:
\[
\begin{array}{cl}
 & \canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= & (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{array}
\]
\noindent 
which places the strings that match $Q_i$ and $Q$ according to the
canonical permutation (i.e., the identity permutation) before applying
the canonizers $\canonize(Q_i)$ and $\canonize(Q)$, respectively.

@@

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_{\normalize{R_1}{R_2}{f}} \; w' &\iff&
f(w)=f(w') \\
w \; \equiv_{\id(R)} \; w' &\iff& w = w' \\
w \; \equiv_{\collapse{R}{s}} \; w' &\iff& \mathit{True} \\ % \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well-Formed QREs}
\label{subsec:well-formed-qres}

@@

Strong unambiguity is necessary in the premises of the regular QRE combinators
because it ensures that the canonizing function of a QRE is well-defined.
For example, consider the QRE \lstinline{"a"* . (collapse "a"* $\to$ "a")}. The
behaviour of this QRE is not well-defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. We also require that the regular
combinators applied to the kernels of QREs be unambiguous since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.

@@

Figure~\ref{fig:qrerules} gives the inference rules for deriving well-formed
QREs.
\begin{figure}[tp!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\RightLabel{(Id)}
\UnaryInfC{$\wf{\kw{id}(R)}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\RightLabel{(Collapse)}
\UnaryInfC{$\wf{\collapse{R}{s}}$}
\DisplayProof
\hskip 1.5em
\end{center}
